\documentclass[10pt,twoside,final]{memoir}

\title{The Group of Octads}
\author{Bhante Varado}
\date{}

\usepackage{mylayout}

% \hyphenation{}

%% PDF/X-3:2002 info
%% =================

\pdfobjcompresslevel=0%
\pdfminorversion=3%
\pdfinfo{
  /Title (The Group of Octads)
  /Author (Bhante Varado)
  /Subject (buddhism)
  /Keywords (buddhism)
  /GTS_PDFXVersion (PDF/X-3:2002)
}%
\pdfpageattr{
/MediaBox [0 0 369.00000 594.00000]
/BleedBox [0.00000 0.00000 369.00000 594.00000]
/CropBox [0 0 369.00000 594.00000]
/TrimBox [0.00000 0.00000 369.00000 594.00000]
}
\pdfcatalog{
  /PageMode /UseNone
  /OutputIntents [ <<
    /Info (none)
    /Type /OutputIntent
    /S /GTS_PDFX
    /OutputConditionIdentifier (Blurb.com)
    /RegistryName (http://www.color.org/)
  >> ]
}%

%% ==============

\makepagenote

\makeoddfoot{normalpage}{\tiny\bfseries TEST PRINT \textbullet\ TRANSLATION TO BE REVISED}{}{\thepage}
\makeevenfoot{normalpage}{\thepage}{}{\tiny\bfseries TEST PRINT \textbullet\ TRANSLATION TO BE REVISED}

\makeoddfoot{octadpage}{\tiny\bfseries TEST PRINT \textbullet\ TRANSLATION TO BE REVISED}{}{\thepage}
\makeevenfoot{octadpage}{\thepage}{}{\tiny\bfseries TEST PRINT \textbullet\ TRANSLATION TO BE REVISED}


\begin{document}

\thispagestyle{empty}
{\centering\par

\vspace*{3\baselineskip}
Sutta Nip\=ata IV
\bigskip

A\d{t}\d{t}haka Vagga
\bigskip

The Group of Octads
\bigskip

Translated by Bhante Varado

}

\cleardoublepage

\chapterstyle{contents}

\tableofcontents*

% label regexp:
% ^([0-9]{3})\.
% \\verseref{\1.}\\label{atthaka-\1}

% \chapter{Preface}
% \input{./src-tex/preface.tex}
% 
% \chapter{Preface v2}
% \input{./src-tex/preface_v2.tex}

% \chapter{Copyright notice}
% \chapter{Dedication}
% \chapter{Bibliography}

% \pagenumbering{arabic}
% \setcounter{page}{1}

\setcounter{chapter}{0}

\chapterstyle{octad}
\pagestyle{octadpage}

\input{./src-tex/sutta_snp-4.tex}

\addtocontents{toc}{\newpage\par APPENDIX \par\vspace*{0.5\baselineskip}}

\setcounter{chapter}{0}

\chapterstyle{appendix}
\pagestyle{normalpage}

\midsloppy

\chapter{The Octads in a Nutshell}
\input{./src-tex/appendix_the-octads-in-a-nutshell.tex}

\chapter{Goal and path: what is the relationship?}\label{goal-and-the-path}
\input{./src-tex/appendix_goal-and-path.tex}

\chapter{Do arahants think `I am'?}
\input{./src-tex/appendix_do-arahants-think-i-am.tex}

\chapter{Do arahants compare themselves?}
\input{./src-tex/appendix_do-arahants-compare.tex}

\chapter{Did the Buddha think Buddhism was the best?}
\input{./src-tex/appendix_did-the-buddha-think-buddhism.tex}

\chapter{What happens to an arahant at death?}
\input{./src-tex/appendix_what-happens-to-an-arahant.tex}

\chapter{Is the arahant free of kamma?}\label{arahant-free-of-kamma}
\input{./src-tex/appendix_is-the-arahant-free-of-kamma.tex}

\chapter{Do arahants blame themselves?}\label{arahant-blame-themselves}
\input{./src-tex/appendix_do-arahants-blame-themselves.tex}

\chapter{What is the consciousness of an arahant?}\label{arahant-consciousness}
\input{./src-tex/appendix_what-is-the-consciousness-of-an-arahant.tex}

\chapter{Should monks forsake precepts and practices?}\label{should-monks-forsake}
\input{./src-tex/appendix_should-monks-forsake-precepts.tex}

% \chapter{Transcending religious differences: a modern perspective}
% \input{./src-tex/appendix_transcending-religious-differences.tex}

\chapter{The Story of Venerable So\d{n}a}
\input{./src-tex/appendix_story-of-ven-sona.tex}

\chapter{Venerable Mah\=akacc\=ana's explanation of verse 844}\label{mahakaccana-v844}
\input{./src-tex/appendix_mahakaccana-explanation-verse-844.tex}

\chapter{Do arahants have differences in opinion?}
\input{./src-tex/appendix_do-arahants-have-differences.tex}

\chapter{Translation notes}
\input{./src-tex/appendix_translation-notes.tex}

\mbox{}\newpage\thispagestyle{empty}\mbox{}

\end{document}
