
Venerable Mahākaccāna was one of the eminent disciples of the Buddha, considered chief amongst monks who could explain the brief sayings of the Buddha. He was born in Ujjenī town (modern-day Ujjain) in a country called Avanti. This country, now part of modern-day Madhya Pradesh, lay 400 kilometres south-west of Sāvatthi, one of the centres of early Buddhism. Having travelled to visit the Buddha and having attained arahantship, Mahākaccāna returned to Avanti.

Mahākaccāna had a lay supporter in Avanti called Soṇa Kuṭikaṇṇa. Soṇa was keen to become a monk but Mahākaccāna discouraged him by recounting the difficulties of the monks' life: `Difficult, Soṇa, for as long as life lasts are solitary sleeping places, eating once a day and the \textit{brahmacariya} (celibacy). Please, Soṇa, remain a householder as you are and on the Observance days practise the solitary sleeping place, the one meal a day and the \textit{brahmacariya}'.
Soṇa was not to be discouraged, however, and eventually Mahākaccāna agreed to ordain him. As Avanti was so far from the centre of Buddhist practice in India, it took three years to gather the necessary quorum of monks for the ceremony.

After spending his first \textit{vassa} (rainy season) in Avanti, Venerable Soṇa decided to visit the Buddha. When he asked for permission, Mahākaccāna told him: `Yes, very good, Soṇa. Go and see the Sublime One, the Arahant, the perfectly and completely awakened one. You will see the Sublime One, lovely to behold, inspiring of confidence, with senses calmed, with tranquil mind, attained to the greatest self-mastery and serenity, who is tamed, watchful, well controlled, a great being. In my name bow your head to his feet and say: ``My preceptor, Venerable Mahākaccāna, bows his head to your feet and asks if you are well, in good health, with little illness, vigorous and abiding in comfort''\thinspace'.

With this, Soṇa set off and eventually arrived at Jeta's Grove, Anāthapiṇḍika's Monastery in Sāvatthi, and paid his respects to the Buddha. The Buddha told Venerable Ānanda to `Assign a place for this visiting monk to rest'. Ānanda realised `The Lord wants to share a dwelling with this visiting monk' and assigned him a place in the Buddha's own dwelling.

The Buddha and Venerable Soṇa spent the early part of that night meditating outside, then rested till early dawn. When the Buddha arose, he invited Soṇa to recite some of the teachings. Fulfilling this request, Soṇa recited the Group of Octads. When he had finished, the Buddha praised him, saying:

`Well done, monk! The Group of Octads is well memorised by you. You have pondered it carefully, reflected upon it thoroughly. You have a beautiful voice, a good delivery, and clear articulation. You made the meaning clear' (Ud.57; Vin.1.193). The Buddha later declared `Chief among my monks of beautiful speech is Soṇa Kuṭikaṇṇa' (AN.1.24).
